Overview

### 2024-01-04 analysis/001_overview.Rmd
playing around with color scales

```{r area_oa_over_the_years}
oa_ta_plot_df <- ta_df |>
  select(cr_year, ta_oa, no_ta) |>
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  inner_join(year_all)
  # plot
oa_ta_plot <- ggplot(oa_ta_plot_df, aes(cr_year, value / n, fill = type, group = type)) +
  geom_area() +
  scale_fill_manual(NULL, values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement"),  guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::percent_format(), limits = c(0, .15), breaks = c(0, .05, .1, .15)) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(legend.position="top", legend.justification = "left", axis.text.x = element_blank())
```

@@

publisher_prop <- ggplot(publisher_summary_df, aes(gsub("^20", "'", as.character(cr_year)), prop, fill = type, group = type)) +
  geom_area() +
  facet_wrap(~publisher, nrow = 1) +
  scale_fill_manual("Hybrid OA", values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement"),  guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::percent_format(),
                     breaks = c(0.05, 0.1, 0.15, 0.2, 0.25)) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(legend.position="top", legend.justification = "left", axis.text.x = element_blank())

@@

```{r subject_plots}
subject_trend_plot <- top_level_by_year_oa |>
  filter(top_level != "Multidisciplinary") |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  select(cr_year, top_level, contains("frac"),-oa_frac) |>
  pivot_longer(
    cols = c(ta_oa_frac, no_ta_fractions),
    names_to = "type",
    values_to = "value"
  ) |>
  mutate(prop = value / articles_frac) |>
  ggplot(aes(
    gsub("^20", "'", as.character(cr_year)),
    prop,
    fill = type,
    group = type
  )) +
  geom_area() +
  facet_wrap( ~ top_level, nrow = 1) +
  scale_fill_manual(
    "Hybrid OA",
    values = c(
      no_ta_fractions = "#b3b3b3a0",
      ta_oa_frac = "#56B4E9"
    ),
    labels = c(no_ta_fractions = "Other", ta_oa_frac = "Agreement"),
    guide = guide_legend(reverse = TRUE)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05)),
    labels = scales::percent_format(),
    breaks = c(0.05, 0.1, 0.15, 0.2, 0.25)
  ) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    axis.text.x = element_blank()
  )

@@

```{r, country_plot, fig.asp=0.8}
ggplot(top_20_plot_df, aes(
    gsub("^20", "'", as.character(cr_year)),
    value / all_total,
    group = type,
    fill = type
  )) +
  geom_area(stat = "identity") +
  scale_fill_manual(
    "Hybrid OA",
    values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"),
    labels = c(no_ta = "Other", ta_oa = "Agreement"),
    guide = guide_legend(reverse = TRUE)
  ) +
  facet_wrap( ~ country_name_fct,
              nrow = 3,
              labeller = as_labeller(country_facet_labels)) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1L),
    expand = expansion(mult = c(0, 0.05)),
    breaks = scales::breaks_extended(4)
  ) +
  theme_minimal() +
  labs(y = "OA Share", x = "2018 - 2022") +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10),
    strip.text = ggtext::element_markdown(size = 6),
    plot.subtitle = ggtext::element_markdown(hjust = 0.5),
    panel.spacing = unit(1, "lines")
  )
```

>>>>

```{r area_oa_over_the_years}
oa_ta_plot_df <- ta_df |>
  select(cr_year, ta_oa, no_ta) |>
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  inner_join(year_all)
  # plot
oa_ta_plot <- ggplot(oa_ta_plot_df, aes(cr_year, value / n, fill = type, group = type)) +
  geom_area(color = "white",
            alpha = 0.8,
            linewidth = 0.3) +
  scale_fill_manual(NULL, values = c(no_ta = "#b3b3b3", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement"),  guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::percent_format(), limits = c(0, .15), breaks = c(0, .05, .1, .15)) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(legend.position="top", legend.justification = "left", axis.text.x = element_blank())
```

@@

publisher_prop <- ggplot(publisher_summary_df, aes(gsub("^20", "'", as.character(cr_year)), prop, fill = type, group = type)) +
    geom_area(color = "white",
            alpha = 0.8,
            linewidth = 0.3) +
  facet_wrap(~publisher, nrow = 1) +
  scale_fill_manual("Hybrid OA", values = c(no_ta = "#b3b3b3", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement"),  guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::percent_format(),
                     breaks = c(0.05, 0.1, 0.15, 0.2, 0.25)) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(legend.position="top", legend.justification = "left", axis.text.x = element_blank())

@@

```{r subject_plots}
subject_trend_plot <- top_level_by_year_oa |>
  filter(top_level != "Multidisciplinary") |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  select(cr_year, top_level, contains("frac"),-oa_frac) |>
  pivot_longer(
    cols = c(ta_oa_frac, no_ta_fractions),
    names_to = "type",
    values_to = "value"
  ) |>
  mutate(prop = value / articles_frac) |>
  ggplot(aes(
    gsub("^20", "'", as.character(cr_year)),
    prop,
    fill = type,
    group = type
  )) +
    geom_area(color = "white",
            alpha = 0.8,
            linewidth = 0.3) +
  facet_wrap( ~ top_level, nrow = 1) +
  scale_fill_manual(
    "Hybrid OA",
    values = c(
      no_ta_fractions = "#b3b3b3",
      ta_oa_frac = "#56B4E9"
    ),
    labels = c(no_ta_fractions = "Other", ta_oa_frac = "Agreement"),
    guide = guide_legend(reverse = TRUE)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05)),
    labels = scales::percent_format(),
    breaks = c(0.05, 0.1, 0.15, 0.2, 0.25)
  ) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    axis.text.x = element_blank()
  )

@@

```{r, country_plot, fig.asp=0.8}
ggplot(top_20_plot_df, aes(
    gsub("^20", "'", as.character(cr_year)),
    value / all_total,
    group = type,
    fill = type
  )) +
  geom_area(color = "white",
            alpha = 0.8,
            linewidth = 0.3,
            stat = "identity") +
scale_fill_manual(
    "Hybrid OA",
    values = c(no_ta = "#b3b3b3", ta_oa = "#56B4E9"),
    labels = c(no_ta = "Other", ta_oa = "Agreement"),
    guide = guide_legend(reverse = TRUE)
  ) +
  facet_wrap( ~ country_name_fct,
              nrow = 3,
              labeller = as_labeller(country_facet_labels)) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1L),
    expand = expansion(mult = c(0, 0.05)),
    breaks = scales::breaks_extended(4)
  ) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = "2018 - 2022") +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    axis.text.x = element_blank(),
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10),
    strip.text = ggtext::element_markdown(size = 6),
    plot.subtitle = ggtext::element_markdown(hjust = 0.5),
    panel.spacing = unit(1, "lines")
  )
```

here a scatterplot and a classifciation of oecd / brics / other. first, the scatterplot 2022!

```{r}
oecd <- readr::read_csv(here::here("data", "oecd.csv")) |>
  mutate(country_code = countrycode::countrycode(iso3, "iso3c", "iso2c"))

country_scatter_df <- country_df |>
#  filter(cr_year == "2022", !is.na(country_code)) |>
  mutate(cat = case_when(
    country_code %in% oecd$country_code ~ "OECD",
    country_code %in% c("BR", "CN", "IN", "RU", "ZA") ~ "BRICS",
    .default = "Other"
  ))

```{r country_bar_chart_top_20}
top_20_df <- country_scatter_df |>
  filter(cr_year == "2022") |>
  mutate(country_name_fct = forcats::fct_other(country_name, keep = top_20$country_name)) |>
  mutate(country_name_fct = forcats::fct_relevel(country_name_fct, top_20$country_name, "Other")) |>
  group_by(country_name_fct, country_code, cat) |>
  summarise(all = sum(all),
            oa = sum(oa),
            ta_oa = sum(ta_oa), .groups = "drop") |>
  head(20) |>
  mutate(cat_labels = case_when(
   cat == "OECD" ~ glue::glue("<span style = 'color: #3fc1c9'>{country_code}</span>"),
   cat == "BRICS" ~ glue::glue("<span style = 'color: #fc5185'>{country_code}</span>"),
   cat == "Other" ~ glue::glue("<span style = 'color: #e1e1e1'>{country_code}</span>")
  )) |>
  mutate(cat_labels = fct_inorder(cat_labels))

top_20_plot <- top_20_df |>
  mutate(no_ta = oa - ta_oa) |>
  pivot_longer(
  cols = c(no_ta, ta_oa),
  values_to = "value",
  names_to = "type") |>
  ggplot(aes(reorder(as.character(country_code),  value), value / sum(value), fill = type)) +
    geom_bar(stat = "identity", alpha = 0.8, color = "#000000", linewidth = 0.2) +
   scale_fill_manual(
    NULL,
    values = c(no_ta = "#b3b3b3", ta_oa = "#56B4E9"),
    labels = c(no_ta = "Other", ta_oa = "Agreement"),
    guide = FALSE) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1L),
    expand = expansion(mult = c(0, 0.05)),
    breaks = scales::breaks_extended(4)
  ) +
  theme_minimal() +
#  theme(axis.text.y = ggtext::element_markdown()) +
  labs(x = NULL, y = "% of Hybrid OA published\nin 2022")

```{r}
country_scatter_22 <- country_scatter_df |>
  filter(cr_year == "2022", !is.na(country_code)) |>
  ggplot(aes(all, oa)) +
  geom_point(aes(fill = cat, size = ta_oa), pch = 21, alpha = 0.7) +
  scale_fill_manual(NULL, values = c("OECD" = "#3fc1c9", "BRICS" = "#fc5185", "Other" = "#e1e1e1")) +
  geom_smooth(aes(all, oa), method=lm) +
  scale_size("TA OA", guide = "none") +
  ggrepel::geom_text_repel(
    data = filter(country_scatter_df, cr_year == 2022, cat %in% c("OECD", "BRICS"), ta_oa > 5000 | country_code %in% c("CN", "IN")),
    aes(x = all, y = oa, label = country_code),
    force = 5,
    segment.color = "grey30",
    segment.size = 0.2,
    segment.alpha = 0.8,
    color = "grey10", 
    size = 3,
    box.padding = 0.3) +
  scale_x_continuous(labels =  scales::number_format(big.mark = ","), guide = guide_axis(angle = 45)) +
  scale_y_continuous(labels =  scales::number_format(big.mark = ",")) +
  theme_minimal() +
  labs(x = "Total articles in hybrid journals in 2022", y = "OA articles") +
  theme(legend.position = "top",
    legend.justification = "left",
    legend.direction = "horizontal")

oecd_stats <- readr::read_csv(here::here("data", "oecd_stats.csv"))
oecd_stats_df <- oecd_stats |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  group_by(cr_year, cat) |>
  summarise(all = sum(articles, na.rm = TRUE), oa = sum(oa,na.rm = TRUE), ta_oa = sum(ta_oa, na.rm = TRUE)) |>
  mutate(closed = all - oa,
         oa_other = oa - ta_oa) |>
  pivot_longer(cols = c(ta_oa, oa_other, closed), names_to = "oa_type") |>
  mutate(cat = ifelse(is.na(cat), "Other", cat))

country_bar <- ggplot(oecd_stats_df, aes(gsub("^20", "'", as.character(cr_year)), value, group = oa_type, fill = oa_type, color = oa_type)) +
  geom_area(color = "#000000",
            alpha = 0.8,
            linewidth = 0.2) +
  scale_color_manual(values = colorspace::darken(c(oa_other = "#b3b3b3", ta_oa = "#56B4E9", closed = "#e1e1e1"), 0.3)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels =  scales::number_format(big.mark = ",")) +
  facet_wrap(~ cat) +
  scale_fill_manual("Hybrid OA", values = c(oa_other = "#b3b3b3", ta_oa = "#56B4E9", closed = "#e1e1e1"),
                    labels = c(oa_other = "Other", ta_oa = "Agreement", closed = "NA"),  guide = guide_legend(reverse = TRUE)) +
  theme_minimal() +
  labs(x = "Publication year", y = "Total articles")

```{r country_patch, fig.asp=0.9}
library(patchwork)
wrap_plots(
  A = country_bar,
  B = country_scatter_22,
  C = top_20_plot,
  design = "
  AC
  BC",
#  guides = "collect",
  widths = c(0.7, 0.3),
  heights = c(0.5,0.5)
) &
  theme(
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10),
    legend.position = "top",
    legend.justification = "left",
    legend.direction = "horizontal") &
    plot_annotation(tag_levels = 'A') 
```
